Weighted Model Counting Without Parameter Variables

### 2021-05-09 doc/paper3/paper.tex
Ready?

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage[binary-units]{siunitx}
\usepackage{booktabs}
\usepackage{capt-of}
\usepackage[inline]{enumitem}

@@

% TODO: Check author % instructions in the lncs/ directory.
% TODO: Make sure that the paper fits into 15 pages.

@@

\author{Paulius Dilkas\inst{1} \and
Vaishak Belle\inst{1,2}}
% TODO: maybe add an envelope or two?

@@

Weighted model counting (WMC), i.e., a generalisation of propositional model
counting that assigns weights to literals and computes the total weight of all
models of a propositional formula \cite{DBLP:journals/ai/ChaviraD08}, has
emerged as a powerful computational framework for problems in many domains,
e.g., probabilistic graphical models such as Bayesian networks and Markov
networks
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05},
neuro-symbolic artificial intelligence \cite{DBLP:conf/icml/XuZFLB18},
probabilistic programs \cite{DBLP:journals/pacmpl/HoltzenBM20}, and
probabilistic logic programs \cite{DBLP:journals/tplp/FierensBRSGTJR15}. It has
been extended to support continuous variables \cite{DBLP:conf/ijcai/BellePB15},
infinite domains \cite{DBLP:conf/aaai/Belle17}, first-order logic
\cite{DBLP:journals/cacm/GogateD16,DBLP:conf/ijcai/BroeckTMDR11}, and arbitrary
semirings \cite{DBLP:journals/ijar/BelleR20,DBLP:journals/japll/KimmigBR17}.
However, as the definition of WMC puts weights on literals, additional variables
often need to be added for the sole purpose of holding a weight
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05}.
As the parameterised complexity of model counting (and, by extension, WMC)
depends on the number of variables
\cite{DBLP:conf/focs/BacchusDP03,DBLP:journals/jda/SamerS10}, this can make WMC
unnecessarily slow and could be detrimental to WMC algorithms such as
\textsf{ADDMC} \cite{DBLP:conf/aaai/DudekPV20} that depend on variable ordering
heuristics.
% TODO: While the complexity results in the cited papers [2, 32] have various
% widths as the dominant (i.e., exponential) part of the formula, the number of
% variables is featured as well, e.g., as $n$ in [2]. We did not notice a
% significant difference in treewidth before and after performing our
% transformation (we can include details about this in the final version of the
% paper). The observed performance improvement is more likely related to the
% variable ordering heuristic used by ADDs.
% One important aspect that is not analyzed is how the technique affects the
% treewidth of the instance.
% This is important because DPMC makes use of tree decompositions of low width.
% In particular, from a theoretical standpoint the questions are whether the
% given reduction can increase or decrease the treewidth, and if yes, are there
% any bounds to it.
% (to me it seems that the reduction can decrease treewidth but cannot increase
% it)
% From practical standpoint the question is how the widths of the tree
% decompositions used by DPMC+bklm16 and DPMC+bklm16++ compare.
% Do you have any data on this?
% Also related to treewidth, the sentence "As the parameterised complexity of
% model counting (and, by extension, WMC) depends on the number of variables [2,
% 32], ..." sounds very weird because the cited papers consider the complexity
% parameterized by treewidth.
% Should "number of variables" be replaced by "treewidth"?

@@

% TODO: We chose to transform existing encodings instead of creating a new one
% because encodings such as cd05, cd06, and bklm16 do a good job at reducing
% each CPT to a minimal logical representation by using techniques such as prime
% implicants and limited forms of resolution. We wanted to reuse and build on
% these ideas.

@@

WMC is a well-established technique for Bayesian network inference, particularly
effective on networks where most variables have only a few possible values
\cite{DBLP:conf/kr/Darwiche02}. Many ways of encoding a Bayesian network into a
WMC instance have been proposed. We will refer to them based on the initials of
the authors and the year of publication. Darwiche was the first to suggest the
\texttt{d02} \cite{DBLP:conf/kr/Darwiche02} encoding that, in many ways, remains
the foundation behind most other encodings. He also introduced the distinction
between \emph{indicator} and \emph{parameter variables}; the former represent
variable-value pairs in the Bayesian network, while the latter are associated
with probabilities in the CPTs. The encoding \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05} is the only encoding that deviates from this
arrangement: for each variable in the Bayesian network, one indicator variable
acts simultaneously as a parameter variable. Chavira and Darwiche propose
\texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05} where they shift from WMC to
minimum-cardinality WMC because that allows the encoding to have fewer variables
and clauses. In particular, they propose a way to use the same parameter
variable to represent all probabilities in a CPT that are equal and keep only
clauses that `imply' parameter variables (i.e., omit clauses where a parameter
variable implies indicator variables).\footnote{\Cref{example:2} demonstrates
  what we mean by implication clauses.} In their next encoding, \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}, the same authors optimise the aforementioned
implication clauses, choosing the smallest sufficient selection of indicator
variables. A decade later, Bart et al. present \texttt{bklm16}
\cite{DBLP:conf/ecai/BartKLM16} that improves upon \texttt{cd06} in two ways.
First, they optimise the number of indicator variables used per Bayesian network
variable from a linear to a logarithmic amount. Second, they introduce a scaling
factor that can `absorb' one probability per Bayesian network variable. However,
for this work, we choose to disable the latter improvement since this scaling
factor is often small enough to be indistinguishable from zero without the use
of arbitrary precision arithmetic, making it completely unusable on realistic
instances. Indeed, the reader is free to check that even a small Bayesian
network with seven mutually independent binary variables, 0.1 and 0.9
probabilities each, is already big enough for the scaling factor to be exactly
equal to zero (as produced by the \texttt{bklm16}
encoder\footnote{\url{http://www.cril.univ-artois.fr/kc/bn2cnf.html}}). We
suspect that this issue was not identified during the original set of
experiments because they never looked at numerical answers.

@@

We introduce a new type of computational problem called \emph{pseudo-Boolean
  projection} based on two-valued pseudo-Boolean functions. While the same
computational framework can handle any pseudo-Boolean functions, two-valued
functions are particularly convenient because \textsf{DPMC} can be easily
adapted to use them as input, and they are easily representable in both text
files and using the syntax proposed in this paper.

@@

% TODO: Examples can be added for illustration of the transformation algorithm.
% TODO:  We will update the camera-ready version of the paper with a complete
% description of how the input format was adjusted to support a new type of
% problem instance. In short, we replace the weight lines of the form 'w
% <variable> <positive weight>' with 'w <conjunction of literals> <positive
% weight> <negative weight>'. The input format is restricted to conjunctions for
% simplicity, but, of course, the algorithm can accept arbitrary formulas.
\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC (or minimum-cardinality WMC) instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X_I, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{$v \in X_I$ such that $\{[v]_1^p, [\neg v]_1^q\}
    \subseteq F$ for some $p$ and $q$ \label{line:foreach2start}}{
    $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
    \}$\; \label{line:foreach2end}
  }
\end{algorithm}

In this section, we describe an algorithm for transforming WMC instances to the
PBP format while removing all parameter variables. The algorithm works on four
out of the five Bayesian network encodings: \texttt{bklm16}
\cite{DBLP:conf/ecai/BartKLM16}, \texttt{cd05}
\cite{DBLP:conf/ijcai/ChaviraD05}, \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}, and \texttt{d02}
\cite{DBLP:conf/kr/Darwiche02}. There is no obvious way to adjust it to work
with \texttt{sbk05} because the roles of indicator and parameter (i.e.,
`chance') variables overlap \cite{DBLP:conf/aaai/SangBK05}. The algorithm is
based on several observations that will be made more precise in
\cref{sec:proof}. First, all weights except for $\{w(p) \mid p \in X_P\}$ are
redundant as they either duplicate an already-defined weight or are equal to
one. Second, each clause has at most one parameter variable. Third, if the
parameter variable is negated, we can ignore the clause (this idea first appears
in the \texttt{cd05} paper \cite{DBLP:conf/ijcai/ChaviraD05}). Note that while
we formulate our algorithm as a sequel to the WMC encoding procedure primarily
because the implementations of Bayesian network WMC encodings are all
closed-source, as all transformations in the algorithm are local, it can be
efficiently incorporated into a WMC encoding algorithm with no slowdown.

The algorithm is listed as \cref{alg:transformation}. The main part of the
algorithms is the first loop that iterates over clauses. If a clause consists of
a single parameter variable, we incorporate it into $\omega$. If a clause is
of the form $\alpha \Rightarrow p$, where $p \in X_P$ and $\alpha$ is a
conjunction of literals over $X_I$, we transform it into a pseudo-Boolean
function $[\alpha]_1^{w(p)}$. If a clause (say, $c \in \phi$) has no parameter
variables, we reformulate it into a pseudo-Boolean function $[c]_0^1$. Finally,
if a clause has negative parameter literals, we skip it.

@@

\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the WMC answer is
  \eqref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that $\sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models
    \phi} \prod_{Y \models l} w(l)$ as required.
\end{proof}

@@

  Let us first consider how the unfinished WMC instance $(F, X_I, \omega)$ after
  the loop on \crefrange{line:foreach1start}{line:foreach1end} differs from
  \eqref{eq:new_wmc2}. Note that \cref{alg:transformation} leaves each $c \in
  \phi_*$ unchanged, i.e., adds $[c]_0^1$ to $F$. We can then fix an arbitrary
  $p \in X_P$ and let $F_p$ be the set of functions added to $F$ as a
  replacement of $\phi_p$. It is sufficient to show that
  \begin{equation} \label{eq:to_show}
    \omega \prod_{f \in F_p} f = \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1.
  \end{equation}
  Note that under \cref{cond:d02}, $\bigwedge_{c \in \phi_p} c \equiv p
  \Leftrightarrow \bigwedge_{i=1}^n l_i$ for some family of indicator variable
  literals $(l_i)_{i=1}^n$. Thus, $\exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in
    \phi_p} [c]_0^1 = \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow
    \bigwedge_{i=1}^n l_i \right]_0^1$. If $w(p) = 1$, then
  \begin{equation} \label{eq:bigsums}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 = \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=0}.
  \end{equation}
  Since for any input, $\bigwedge_{i=1}^n l_i$ is either true or false, exactly
  one of the two summands in \cref{eq:bigsums} will be equal to one, and the
  other will be equal to zero, and so
  \[
    \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
      \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
        l_i \right]_0^1\right|_{p=0} = 1,
  \]
  where $1$ is a pseudo-Boolean function that always returns one. On the other
  side of \cref{eq:to_show}, since $F_p = \emptyset$, and $\omega$ is unchanged,
  we get $\omega\prod_{f \in F_p} f = 1$, and so \cref{eq:to_show} is satisfied
  under \cref{cond:d02} when $w(p) = 1$.

  If $w(p) \ne 1$, then $F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i
    \right]_1^{w(p)} \right\}$, and $\omega = 1$, and so we want to show that
  $\left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
  \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1$, and indeed
  \[
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[
      \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i
    \right]_1^{w(p)}.
  \]
  This finishes the proof of the correctness of the first loop under
  \cref{cond:d02}.

  Now let us assume \cref{cond:bklm16}. We still want to prove
  \cref{eq:to_show}. If $w(p) = 1$, then $F_p = \emptyset$, and $\omega = 1$,
  and so the left-hand side of \cref{eq:to_show} is equal to one. Then the
  right-hand side is
  \[
    \exists_p [p]_0^1 \prod_{c \in \phi_p} [c]_0^1 = \exists_p \left[ p \land
      \bigwedge_{c \in \phi_p} c \right]_0^1 = \exists_p [p]_0^1 = 0 + 1 = 1
  \]
  since $p \in c$ for every clause $c \in \phi_p$.

  If $w(p) \ne 1$, and $\{ p \} \in \phi_p$, then, by
  \cref{cond:just_parameter}, $\phi_p = \{ \{ p \} \}$, and
  \cref{alg:transformation} produces $F_p = \emptyset$ and $\omega = w(p)$, and
  so
  \[
    \exists_p [p]_{w(\neg p)}^{w(p)} [p]_0^1 = \exists_p [p]^{w(p)}_0 = w(p) =
    \omega \prod_{f \in F_p} f.
  \]
  The only remaining case is when $w(p) \ne 1$ and $\{ p \} \not \in \phi_p$.
  Then $\omega = 1$, and $F_p = \left\{ \left[\bigwedge_{l \in c \setminus \{ p
        \}} \neg l\right]_1^{w(p)} \;\middle|\; c \in \phi_p \right\}$, so we
  need to show that $\prod_{c \in \phi_p} \left[\bigwedge_{l \in c \setminus \{
      p \}} \neg l\right]_1^{w(p)} = \exists_p [p]_{1-w(p)}^{w(p)} \prod_{c \in
    \phi_p} [c]_0^1$. We can rearrange the right-hand side as
  \begin{align*}
    \exists_p [p]_{1-w(p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 &= \exists_p [p]_{1-w(p)}^{w(p)} \left[ p \lor \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_0^1 \\
                                                               &= w(p) + (1-w(p)) \left[ \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_0^1 \\
                                                               &= \left[ \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_{w(p)}^1 = \left[ \bigvee_{c \in \phi_p} \bigwedge_{l \in c \setminus \{ p \}} \neg l \right]_1^{w(p)}.
  \end{align*}
   By \cref{cond:disjoint}, $\bigwedge_{l \in c \setminus \{ p \}} \neg l$ can
   be true for at most one $c \in \phi_p$, and so $\left[ \bigvee_{c \in \phi_p}
     \bigwedge_{l \in c \setminus \{ p \}} \neg l \right]_1^{w(p)} = \prod_{c
     \in \phi_p} \left[ \bigwedge_{l \in c \setminus \{ p \}} \neg l
   \right]_1^{w(p)}$ which is exactly what we needed to show. This ends the
   proof that the first loop of \cref{alg:transformation} preserves the answer
   under both \cref{cond:d02} and \cref{cond:bklm16}. Finally, the loop on
   \crefrange{line:foreach2start}{line:foreach2end} of \cref{alg:transformation}
   replaces $[v]_1^p[\neg v]_1^q$ with $[v]_q^p$ (for some $v \in X_I$ and $p, q
   \in \mathbb{R}$), but, of course, $[v]_1^p[\neg v]_1^q = [v]_1^p[v]_q^1 =
   [v]_q^p$, i.e., the answer is unchanged.
\end{proof}

>>>>

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage[binary-units]{siunitx}
\usepackage{booktabs}
\usepackage{capt-of}
\usepackage[inline]{enumitem}
\usepackage[misc,geometry]{ifsym}

@@

@@

\author{Paulius Dilkas\inst{1}$^{(\textrm{\Letter})}$ \and
Vaishak Belle\inst{1,2}}

@@

Weighted model counting (WMC), i.e., a generalisation of propositional model
counting that assigns weights to literals and computes the total weight of all
models of a propositional formula \cite{DBLP:journals/ai/ChaviraD08}, has
emerged as a powerful computational framework for problems in many domains,
e.g., probabilistic graphical models such as Bayesian networks and Markov
networks
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05},
neuro-symbolic artificial intelligence \cite{DBLP:conf/icml/XuZFLB18},
probabilistic programs \cite{DBLP:journals/pacmpl/HoltzenBM20}, and
probabilistic logic programs \cite{DBLP:journals/tplp/FierensBRSGTJR15}. It has
been extended to support continuous variables \cite{DBLP:conf/ijcai/BellePB15},
infinite domains \cite{DBLP:conf/aaai/Belle17}, first-order logic
\cite{DBLP:journals/cacm/GogateD16,DBLP:conf/ijcai/BroeckTMDR11}, and arbitrary
semirings \cite{DBLP:journals/ijar/BelleR20,DBLP:journals/japll/KimmigBR17}.
However, as the definition of WMC puts weights on literals, additional variables
often need to be added for the sole purpose of holding a weight
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05}.
This can be particularly detrimental to WMC algorithms that rely on variable
ordering heuristics.
% As the parameterised complexity of model counting (and, by extension, WMC)
% depends on the number of variables
% \cite{DBLP:conf/focs/BacchusDP03,DBLP:journals/jda/SamerS10}, this can make WMC
% unnecessarily slow and could be detrimental to WMC algorithms such as
% \textsf{ADDMC} \cite{DBLP:conf/aaai/DudekPV20} that depend on variable ordering
% heuristics.

@@

@@

WMC is a well-established technique for Bayesian network inference, particularly
effective on networks where most variables have only a few possible values
\cite{DBLP:conf/kr/Darwiche02}. Many ways of encoding a Bayesian network into a
WMC instance have been proposed. We will refer to them based on the initials of
the authors and the year of publication. Darwiche was the first to suggest the
\texttt{d02} \cite{DBLP:conf/kr/Darwiche02} encoding that, in many ways, remains
the foundation behind most other encodings. He also introduced the distinction
between \emph{indicator} and \emph{parameter variables}; the former represent
variable-value pairs in the Bayesian network, while the latter are associated
with probabilities in the CPTs. The encoding \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05} is the only encoding that deviates from this
arrangement: for each variable in the Bayesian network, one indicator variable
acts simultaneously as a parameter variable. Chavira and Darwiche propose
\texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05} where they shift from WMC to
minimum-cardinality WMC because that allows the encoding to have fewer variables
and clauses. In particular, they propose a way to use the same parameter
variable to represent all probabilities in a CPT that are equal and keep only
clauses that `imply' parameter variables (i.e., omit clauses where a parameter
variable implies indicator variables).\footnote{\Cref{example:2} demonstrates
  what we mean by implication clauses.} In their next encoding, \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}, the same authors optimise the aforementioned
implication clauses, choosing the smallest sufficient selection of indicator
variables. A decade later, Bart et al. present \texttt{bklm16}
\cite{DBLP:conf/ecai/BartKLM16} that improves upon \texttt{cd06} in two ways.
First, they optimise the number of indicator variables used per Bayesian network
variable from a linear to a logarithmic amount. Second, they introduce a scaling
factor that can `absorb' one probability per Bayesian network variable. However,
for this work, we choose to disable the latter improvement since this scaling
factor is often small enough to be indistinguishable from zero without the use
of arbitrary precision arithmetic, making it completely unusable on realistic
instances. Indeed, the reader is free to check that even a small Bayesian
network with seven mutually independent binary variables, 0.1 and 0.9
probabilities each, is already big enough for the scaling factor to be exactly
equal to zero (as produced by the \texttt{bklm16}
encoder\footnote{\url{http://www.cril.univ-artois.fr/kc/bn2cnf.html}}). We
suspect that this issue was not identified during the original set of
experiments because the authors never looked at numerical answers.

@@

We introduce a new type of computational problem called \emph{pseudo-Boolean
  projection} based on two-valued pseudo-Boolean functions. While the same
computational framework can handle any pseudo-Boolean functions, two-valued
functions are particularly convenient because \textsf{DPMC} can be easily
adapted to use them as input. Since we will only encounter functions of
the form $[\phi]^a_b$, where $\phi$ is a conjunction of literals, we can
represent it in text as \texttt{w $\langle\phi\rangle$ a b} where
$\langle\phi\rangle$ is a representation of $\phi$ analogous to the
representation of a clause in the DIMACS CNF format.

@@

\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC (or minimum-cardinality WMC) instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X_I, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{$v \in X_I$ such that $\{[v]_1^p, [\neg v]_1^q\}
    \subseteq F$ for some $p$ and $q$ \label{line:foreach2start}}{
    $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
    \}$\; \label{line:foreach2end}
  }
\end{algorithm}

In this section, we describe an algorithm for transforming WMC instances to the
PBP format while removing all parameter variables. We chose to transform
existing encodings instead of creating a new one to reuse already-existing
techniques for encoding each CPT to its minimal logical representation such as
prime implicants and limited forms of resolution
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}.
The transformation algorithm works on four out of the five Bayesian network
encodings: \texttt{bklm16} \cite{DBLP:conf/ecai/BartKLM16}, \texttt{cd05}
\cite{DBLP:conf/ijcai/ChaviraD05}, \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}, and \texttt{d02}
\cite{DBLP:conf/kr/Darwiche02}. There is no obvious way to adjust it to work
with \texttt{sbk05} because the roles of indicator and parameter variables
overlap \cite{DBLP:conf/aaai/SangBK05}.

The algorithm is based on several observations that will be made more precise in
\cref{sec:proof}. First, all weights except for $\{w(p) \mid p \in X_P\}$ are
redundant as they either duplicate an already-defined weight or are equal to
one. Second, each clause has at most one parameter variable. Third, if the
parameter variable is negated, we can ignore the clause (this idea first appears
in the \texttt{cd05} paper \cite{DBLP:conf/ijcai/ChaviraD05}). Note that while
we formulate our algorithm as a sequel to the WMC encoding procedure primarily
because the implementations of Bayesian network WMC encodings are all
closed-source, as all transformations in the algorithm are local, it can be
efficiently incorporated into a WMC encoding algorithm with no slowdown.

The algorithm is listed as \cref{alg:transformation}. The main part of the
algorithms is the first loop that iterates over clauses. If a clause consists of
a single parameter variable, we incorporate it into $\omega$. If a clause is
of the form $\alpha \Rightarrow p$, where $p \in X_P$, and $\alpha$ is a
conjunction of literals over $X_I$, we transform it into a pseudo-Boolean
function $[\alpha]_1^{w(p)}$. If a clause $c \in \phi$ has no parameter
variables, we reformulate it into a pseudo-Boolean function $[c]_0^1$. Finally,
clauses with negative parameter literals are omitted.

@@

\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the WMC answer of \eqref{eq:new_wmc} is
  $(\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
  (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)$ by \cref{lemma:sum}. Note
  that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that $\sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models
    \phi} \prod_{Y \models l} w(l)$ as required.
\end{proof}

@@

  Let us first consider how the unfinished WMC instance $(F, X_I, \omega)$ after
  the loop on \crefrange{line:foreach1start}{line:foreach1end} differs from
  \eqref{eq:new_wmc2}. Note that \cref{alg:transformation} leaves each $c \in
  \phi_*$ unchanged, i.e., adds $[c]_0^1$ to $F$. We can then fix an arbitrary
  $p \in X_P$ and let $F_p$ be the set of functions added to $F$ as a
  replacement of $\phi_p$. It is sufficient to show that
  \begin{equation} \label{eq:to_show}
    \omega \prod_{f \in F_p} f = \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1.
  \end{equation}
  Note that under \cref{cond:d02}, $\bigwedge_{c \in \phi_p} c \equiv p
  \Leftrightarrow \bigwedge_{i=1}^n l_i$ for some family of indicator variable
  literals $(l_i)_{i=1}^n$. Thus, $\exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in
    \phi_p} [c]_0^1 = \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow
    \bigwedge_{i=1}^n l_i \right]_0^1$. If $w(p) = 1$, then
  \begin{equation} \label{eq:bigsums}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 = \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=0}.
  \end{equation}
  Since for any input, $\bigwedge_{i=1}^n l_i$ is either true or false, exactly
  one of the two summands in \cref{eq:bigsums} will be equal to one, and the
  other will be equal to zero, and so $\left.\left[ p \Leftrightarrow
      \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=1} + \left.\left[ p
      \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=0} = 1$, where
  $1$ is a pseudo-Boolean function that always returns one. On the other side of
  \cref{eq:to_show}, since $F_p = \emptyset$, and $\omega$ is unchanged, we get
  $\omega\prod_{f \in F_p} f = 1$, and so \cref{eq:to_show} is satisfied under
  \cref{cond:d02} when $w(p) = 1$.

  If $w(p) \ne 1$, then $F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i
    \right]_1^{w(p)} \right\}$, and $\omega = 1$, and so we want to show that
  $\left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
  \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1$. Indeed,
  $\exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
  \right]_0^1 = w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[
    \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i
  \right]_1^{w(p)}$. This finishes the proof of the correctness of the first
  loop under \cref{cond:d02}.

  Now let us assume \cref{cond:bklm16}. We still want to prove
  \cref{eq:to_show}. If $w(p) = 1$, then $F_p = \emptyset$, and $\omega = 1$,
  and so the left-hand side of \cref{eq:to_show} is equal to one. Then the
  right-hand side is $\exists_p [p]_0^1 \prod_{c \in \phi_p} [c]_0^1 = \exists_p
  \left[ p \land \bigwedge_{c \in \phi_p} c \right]_0^1 = \exists_p [p]_0^1 = 0
  + 1 = 1$ since $p \in c$ for every clause $c \in \phi_p$.

  If $w(p) \ne 1$, and $\{ p \} \in \phi_p$, then, by
  \cref{cond:just_parameter}, $\phi_p = \{ \{ p \} \}$, and
  \cref{alg:transformation} produces $F_p = \emptyset$, and $\omega = w(p)$, and
  so $\exists_p [p]_{w(\neg p)}^{w(p)} [p]_0^1 = \exists_p [p]^{w(p)}_0 = w(p) =
  \omega \prod_{f \in F_p} f$. The only remaining case is when $w(p) \ne 1$ and
  $\{ p \} \not \in \phi_p$. Then $\omega = 1$, and $F_p = \left\{
    \left[\bigwedge_{l \in c \setminus \{ p \}} \neg l\right]_1^{w(p)}
    \;\middle|\; c \in \phi_p \right\}$, so we need to show that $\prod_{c \in
    \phi_p} \left[\bigwedge_{l \in c \setminus \{p \}} \neg l\right]_1^{w(p)} =
  \exists_p [p]_{1-w(p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1$. We can rearrange
  the right-hand side as
  \begin{align*}
    \exists_p [p]_{1-w(p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 &= \exists_p [p]_{1-w(p)}^{w(p)} \left[ p \lor \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_0^1 \\
                                                               &= w(p) + (1-w(p)) \left[ \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_0^1 \\
                                                               &= \left[ \bigwedge_{c \in \phi_p} c \setminus \{ p \} \right]_{w(p)}^1 = \left[ \bigvee_{c \in \phi_p} \bigwedge_{l \in c \setminus \{ p \}} \neg l \right]_1^{w(p)}.
  \end{align*}
   By \cref{cond:disjoint}, $\bigwedge_{l \in c \setminus \{ p \}} \neg l$ can
   be true for at most one $c \in \phi_p$, and so $\left[ \bigvee_{c \in \phi_p}
     \bigwedge_{l \in c \setminus \{ p \}} \neg l \right]_1^{w(p)} = \prod_{c
     \in \phi_p} \left[ \bigwedge_{l \in c \setminus \{ p \}} \neg l
   \right]_1^{w(p)}$ which is exactly what we needed to show. This ends the
   proof that the first loop of \cref{alg:transformation} preserves the answer
   under both \cref{cond:d02} and \cref{cond:bklm16}. Finally, the loop on
   \crefrange{line:foreach2start}{line:foreach2end} of \cref{alg:transformation}
   replaces $[v]_1^p[\neg v]_1^q$ with $[v]_q^p$ (for some $v \in X_I$ and $p, q
   \in \mathbb{R}$), but, of course, $[v]_1^p[\neg v]_1^q = [v]_1^p[v]_q^1 =
   [v]_q^p$, i.e., the answer is unchanged.
\end{proof}
